### 2014-10-27 intro.tex
Intro and Motivation
Motivation version 2

Many real-world web applications --- such as Amazon, Facebook, Google, Twitter
--- replicate application state and logic across multiple \emph{replicas} within
and across data centers. Replication is intended not only to improve application
throughput and reduce user perceived latency, but also to tolerate partial
failures without compromising the service availability. Traditionally
programmers have relied on \emph{strong consistency} guarantees, such as
Linerizability~\cite{Herlihy1990} or serializability~\cite{Serializability},
provided by the underlying data store in order to build correct applications.
While strong consistency is an easily stated property, it
masks the reality underlying large-scale distributed systems with respect to
non-uniform latency (i.e., availability) and network
partitions~\cite{Brewer2000,Gilbert}. Indeed, modern web applications, which aim to provide
an "always on" experience, overwhelmingly embrace availability and partition
tolerance at the expense of strong consistency.

Nevetheless, to avoid exposing end user to absurdities such as negative balances
in a bank account, or an item appearing in the shopping cart after it had been
removed~\cite{Dynamo}, web applications \emph{do} require some level of
consistency guarantees from the underlying data store. Accordingly, many
commercial~\cite{} and research~\cite{} systems provide guarantees that are
weaker than strong consistency, such as eventual consistency, causal
consistency, session guarantees, and time line consistency. Each such \emph{weak
consistency} guarantee imposes certain amount of penalty on the availability
(hence, latency and throughput) of the system depending on the extent of
inter-replica coordination required to ensure the guaranteed level of
consistency. It is therefore incumbent on the application programmer to
carefully reason about concurrent conflicting updates, and select the
appropriate store-level weak consistency guarantee sufficient for preserving
application-level integrity properties, while minimizing the adverse impact on
the availability of the application. Understandably, this is an error prone
process as the user not only has to understand the intricate behavior of the
application, but also the semantics of the underlying store consistency levels,
which are often presented informally~\cite{}.

%--KC--
% The choice of correct consistency level is critical to the correctness and
% scalability of the application. Choosing a weaker consistency level than
% required introduces anomalies, while choosing a stronger than necessary level
% affects program scalability.
\KC{Do we include the LWW-Register number here to add emphasis to the fact that
choosing the correct consistency level is essential for program scalability}.

Moreover, weak consistency hinders compositional reasoning about programs.
While an application might be naturally expressed in terms of well-understood
abstract data types, such as dictionaries, counters, queues, and graphs,
geo-distributed stores typically only provide a minimal set of data types with
in-built conflict resolution strategy such as last-writer-wins (LWW) registers,
counters and sets~\cite{Cassandra,DynamoDB}.  Furthermore, while traditional
database systems enable composability through transactions, existing
geo-distributed stores lack unrestricted transactional access to the data.
Working with geo-distributed stores thus requires application state to be
suitably coerced to work within the capabilities of the store.

To address these issues, we introduce \name, a declarative programming system
for eventually consistent data stores. \name's programming model is built around
operation-based Commutative Replicated Data Types (CRDTs) ~\cite{shapiroCRDT}.
The state of an abstract data type in \name is modeled as a replicated
\emph{append-only} log storing the history of \emph{update} operations performed
on the data type . For instance, the state of a replicated counter ADT is stored
on all replicas as a log of \emph{increment} and \emph{decrement} operations
performed on the counter. An operation that indends to read the current value
the counter at a replica only has to \emph{fold} over the history log at that
replica. Our key observation is that under this model, consistency requirements
of a replicated data type operation (eg: \emph{decrement}) can be expressed
effectively in terms of assertions on \emph{membership} property of logs at
various replicas. \name makes use of this observation to introduce a simple
\emph{contract language} that is nevertheless capable of expressing fine-grained
application-level consistency properties.  The contract language is made up of a
couple of primitive relations, namely \emph{visibility} and \emph{session
order}, and a carefully choosen set of logical operators. \name expects
application programmer to use the contract language to describe the set of legal
executions allowed over the replicated data types.  
% The executions descriptions are similar to the high-level axiomatic
% description of relaxed memory models~\cite{Demange2013}.
Subsequently, a contract enforcement system logically analyses contracts using
an SMT solver~\cite{z3} and automatically tags each operation with the weakest
store-level consistency guarantee that is required to satisfy application-level
consistency requirements. In addition, the system statically reports an error if
the given application-level consistency property cannot be satisfied by the
store.

@@

\item \name is implemented as a shim layer over (and importantly without
  modifying) Cassandra, a well-known general-purpose data store.  An
  extensive experimental evaluation over a collection of real-world
  applications, including a Twitter-like micro-blogging site and an
  eBay-like auction site illustrates the practicality of our approach.

>>>>

Many real-world web services --- such as those built and maintained by Amazon,
Facebook, Google, Twitter, etc. --- replicate application state and logic
across multiple \emph{replicas} within and across data centers. Replication is
intended not only to improve application throughput and reduce user-perceived
latency, but also to tolerate partial failures without compromising overall
service availability. Traditionally programmers have relied on \emph{strong
consistency} guarantees such as linerarizability~\cite{Herlihy1990} or
serializability~\cite{Serializability} in order to build correct applications.
While strong consistency is an easily stated property, it masks the reality
underlying large-scale distributed systems with respect to non-uniform latency
(i.e., availability) and network partitions~\cite{Brewer2000,Gilbert}. Indeed,
modern web services, which aim to provide an "always on" experience,
overwhelmingly favour availability and partition tolerance over strong
consistency. To this end, several \emph{weak consistency} models such as
eventual consistency, causal consistency, session guarantees, and timeline
consistency have consequently been proposed.

Under weak consistency, the developer needs to be aware of concurrent
conflicting updates, and has to pay careful attention to avoid unwanted
inconsistencies (e.g., negative balances in a bank account, or having an item
appear in a shopping cart after it has been removed~\cite{Dynamo}). Oftentimes,
the inconsistency leaks from the application and is witnessed by the user.
Ultimately, the developer must decide the consistency level appropriate for a
particular operation; this is understandably an error-prone process requiring
intricate knowledge of both the application as well as the semantics and
implementation of the underlying data store.  More often than not, only
informal descriptions are available~\cite{}.  Picking the correct consistency
level is critical not only for correctness but also for scalability of the
application. While choosing a weaker consistency level than required may
introduce program errors and anomalies, choosing a stronger one than necessary
level can negatively impact program scalability.

%--KC--
% The choice of correct consistency level is critical to the correctness and
% scalability of the application. Choosing a weaker consistency level than
% required introduces anomalies, while choosing a stronger than necessary level
% affects program scalability.
%\KC{Do we include the LWW-Register number here to add emphasis to the fact that
%choosing the correct consistency level is essential for program scalability}.

Moreover, weak consistency hinders compositional reasoning about programs.
While an application might be naturally expressed in terms of well-understood
and expressive data types such as maps, trees, queues, or graphs,
geo-distributed stores typically only provide a minimal set of data types with
in-built conflict resolution strategies such as last-writer-wins (LWW)
registers, counters, and sets~\cite{Cassandra,DynamoDB}.  Furthermor, while
traditional database systems enable composability through transactions,
existing geo-distributed stores lack unrestricted transactional access to the
data.  Working with geo-distributed stores thus requires application state to
be suitably coerced to work within the capabilities of the store.

To address these issues, we describe \name, a declarative programming model and
implementation for eventually consistent stores. The key novelty of \name is an
expressive \emph{contract language} to declare and verify fine-grained
application-level consistency properties. The programmer uses the contract
language to axiomatically describe the set of legal executions allowed over the
replicated data type. Contracts are constructed using primitive consistency
relations such as \emph{visibility} and \emph{session order} along with
standard logical and relational operators. A \emph{contract enforcement system}
automatically maps each operation to a particular consistency level available
on the store. In addition, the system statically reports an error if the given
application-level consistency property cannot be satisfied by the store.

@@

\item \name is implemented as a shim layer over (and importantly, without
	modifying) Cassandra~\cite{Cassandra}, a well-known general-purpose data
	store.  An extensive experimental evaluation over a collection of real-world
	applications, including a Twitter-like micro-blogging site and an eBay-like
	auction site illustrates the practicality of our approach.

### 2014-10-27 motivation.tex
Motivation version 2

In this section, with help of a bank account example, we motivate the
need for a framework to express and enforce fine-grained
application-level consistency requirements under weak consistency
models. 
% \GK{The following text needs to be polished}

% We assume a system model where the eventually consistent database is organized
% as a collection of key-indexed, object-valued tables. The objects stored in the
% tables are instances of the same replicated data type. The database is composed
% of a number of \emph{replicas}, and each object is fully replicated across all
% the replicas. Under this model, a client request for an operation with no
% application-level consistency requirement can be serviced if at least one of
% the replicas is reachable. Thus, with the addition of new replicas, and placing
% them close to the clients, leads to improved throughput and reduced latencies.

\subsection{Bank account database}

Suppose our goal is to implement a highly available bank account
service on top of an eventually consistent data store. For the sake of
presentation, we restrict ourselves to bank account objects (records)
with following simple schema:

\begin{codesql}
Create Table BankAccount (
  accountNo int PRIMARY KEY,
  balance float CHECK (balance >= 0))
\end{codesql}

\noindent and with support for the following \emph{operations}:

\begin{codehaskell}
type AccountNo = Int
type Amount = Float
getBalance :: AccountNo -> Amount
deposit    :: AccountNo -> Amount -> ()
-- returns false if the account has insufficient balance
withdraw   :: AccountNo -> Amount -> Bool
\end{codehaskell}

\begin{figure}[t]
\centering
\subfigure[Unsafe Withdraw]{\label{fig:unsafeWithdrawAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation4}}
\hfill
\subfigure[Negative Balance]{\label{fig:negativeBalanceAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation2}}
\hfill
\subfigure[Missing update]{\label{fig:missingUpdateAnomaly}\includegraphics[width=0.26\columnwidth]{Figures/Motivation1}}
%\hfill
%\subfigure[Monotonicity violation]{\label{fig:monotonicityAnomaly}\includegraphics[width=0.36\columnwidth]{Figures/Motivation3}}
\caption{Anomalies possible under eventual consistency for the get balance operation.}
\label{fig:cleanliness_examples}
\end{figure}

The sole attribute of a \cf{BankAccount} object, identified by its unique
\cf{accountNo}, is its current \cf{balance}. A \cf{deposit} operation
on a \cf{BankAccount} adds some \cf{Amount} to its \cf{balance},
whereas a \cf{withdraw} operation subtracts. Operation \cf{getBalance}
returns the current \cf{balance} in the account. A \cf{withdraw}
operation succeeds and returns \cf{True} only if there is sufficient
balance in the account. Otherwise, it fails and returns \cf{False}.

If we were to use a traditional database system to store
\cf{BankAccount} objects, all that needs to be done is to define the
\cf{BankAccount} schema, as shown above. The database management
system effectively sequentializes all accesses to the database, and
automatically enforces the required integerity constraints on the
data, so that it is never possible for a \cf{getBalance} operation to
return a negative balance, or for a \cf{withdraw} to succeed without
having sufficient balance in the account. However, under eventual
consistency, such anomalies are bound to occur as demonstrated in the
following subsection.

\subsection{Anomalies}

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly}.
Assume that all operations in the figure are on the same
\cf{BankAccount} object. Execution contains two concurrent client
sessions, where one of the sessions executes a \cf{withdraw} followed
by a \cf{getBalance} operation. We say that \c{withdraw} precedes
\cf{getBalance} in \emph{session order}, and visualize it as a
directed \cf{so} edge from the former to the later. Edges labelled
\cf{vis} in the figure capture the \emph{visibility} relation between
operations; an operation is said to be \emph{visible} to other
operation if the later can witness the effects of the former. Two
operations that are not related either via \cf{vis} or \cf{so} are
said to be concurrent. The figure shows an execution where concurrent
\cf{withdraw}s in different sessions succeed independently, but a
subsequent \cf{getBalance} operation that witnesses both the
\cf{withdraw}s reports a negative balance.

It is easy to see that the only way to ensure that integrity
constraint on \cf{balance} is to prevent concurrent withdrawls on the
\cf{BankAccount}. This can be done by insisting that \cf{withdraw} be
executed with strong consistency. Although this prevents unsafe
withdrawls from going through, a strongly consistent \cf{withdraw}
itself is not sufficient to prevent \cf{getBalance} from ever
reporting a negative balance to the end user.

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly},
which consists of three concurrent sessions performing a \cf{deposit},
a \cf{withdraw}, and a \cf{getBalance}, respectively, on the same
\cf{BankAccount}. Assume that initial balance in the account is zero.
As the \cf{vis} edge indicates, operation \cf{withdraw(50)} in session
2, can witnesses the effects of \cf{deposit(100)} from session 1,
leading it to rightfully conclude that the current balance is 100.
Consequently, \cf{withdraw(50)} is performed safely. The effect of
this \cf{withdraw} operation subsequently becomes visible to the
\cf{getBalance} in Session 3. However, \cf{getBalance}, which can only
see the effect of \cf{withdraw} from session 2, but not \cf{deposit}
from session 1, reports the balance of negative 50 to the user.

Negative balance is not the only inconsistency that the user is
exposed to under an eventually consistent bank account. Figure
~\ref{fig:missingUpdateAnomaly} shows an execution where
\cf{getBalance} operation in a session does not witness the effects of
\cf{withdraw} operation performed in the same session, possibly
because \cf{getBalance} was served by a replica that hasn't yet merged
the \cf{withdraw} effect, which could have been served by a different
replica. This anomaly leads the user to incorrectly conclude that the
\cf{withdraw} operation failed to go through.

Although it is easy to understand the reasons behind the occurance of
the aforementioned negative balance and missing update anomalies, the
fix is not quite apparent. Making \cf{getBalance} a strongly
consistent operation is definitely sufficient to avert anomalies, but
is it really necessary? Given that a strongly consistent operation is
significantly expensive, both in terms of latency and monetary value, we ideally
would not want to request strong consistency, unless there are
no alternatives available. Unfortunately, this requires us to
understand subtle differences in semantics of various kinds of weak
consistency alternatives, such as timeline consistency and session
guarantees, offered by off-the-shelf eventually consistent data
stores. This process is error-prone, and becomes further complicated
in presence of transactions, as demonstrated in Sec.
~\ref{sec:transactions}.

\subsection{Implementation in Quelea}

\name automates this error-prone task by letting the programmer
declare application-level consistency constraints as \emph{contracts},
obviating the need to understand store-level consistency guarantees,
and relieving programmer from having to match application
requirements with appropriate store-level alternatives. The
cornerstone of declarative reasoning in \name is the programming model
built around operation-based CRDTs~\cite{shapiroCRDT}. A replicated
data type in \name is defined as a tagged union type, where each tag
is the name of a unique operation defined on the data type. 
%types in Quelea are defined as reductions over the set of effects.
%The following data type definitions capture the operations and effects
%on the bank account and user index objects.
For example, following is the definition of the \cf{BankAccount}
data type in \name:
\begin{codehaskell}
data BankAccount =  Deposit Float | Withdraw Float 
                 | GetBalance
\end{codehaskell}

In Quelea, an object state is nothing but the set of effects (called
the \emph{context}) on this object. As such, \cf{Deposit} and
\cf{Withdraw} effects simply include the amount deposited or
withdrawn, and \cf{AddUser} effect includes the user name. Similarly,
on the user index table, where the key is user name, the value is
simply an effect (\cf{AddName}) binding some user id at this key.
\cf{GetBalance} and \cf{GetName} correspond to the corresponding
read-only operations. The complete definition of the operations on
bank account and user index objects is given below.

\begin{codehaskell}
-- Get balance takes no arguments and read-only
getBalance :: [BankAcc] {- context -} -> () {- args -}
  -> (() {- ret val -}, Maybe BankAcc {- effect -})
getBalance ctxt _ = (sum [v | Deposit v (- ctxt]
        - sum [v | Withdraw v (- ctxt], Nothing)

-- withdraw returns True on success
withdraw :: [BankAcc] -> Float -> (Bool, Maybe BankAcc)
withdraw c v = if sel1 $ getBalance c () >= v
               then (True, Just $ Withdraw v)
               else (False, Nothing)

deposit :: [BankAcc] -> v -> ((), Maybe BankAcc)
deposit _ v = ((), Just $ Deposit v)

addUser :: [BankAcc] -> UserName -> ((), Maybe BankAcc)
addUser _ n = ((), Just $ AddUser n)

addName :: [UserIdx] -> UserID -> (Bool, Maybe UserIdx)
addName [] uid = (True, Just $ AddName uid)
addName x:_ _ = (False, Nothing)

getName :: [UserIdx] -> ()
        -> (Maybe UserName, Maybe UserIdx)
getName [] _ = (Nothing, Nothing)
getName (AddUser uid:_) _ = (Just uid, Nothing)
\end{codehaskell}

The definitions are a straight forward encoding of the expected behavior.
Quelea programming model provides the programmer complete freedom regarding the
semantics and desired convergence property of the replicated data type, and
indeed, we can encode the well-known CRDTs~\cite{SSS} in a declarative fashion.
Observe that the operation definitions presented here only capture the
convergence properties of the replicated data type, and are not concerned with
the consistency properties, which is expressed through the contract language.

@@

The contract for the strongly consistent \cf{addName} operation is given below:
\begin{smathpar}
\begin{array}{l}
\rsf{addNameCtrt} = \forall (a:\rsf{AddName}). \\
\qquad \sameobj{a}{\cureff} \Rightarrow \vis{a}{\cureff} \vee \vis{\cureff}{a} \vee a = \cureff
\end{array}
\end{smathpar}

In the above definition, $\cureff$ represents the current effect --- the effect
emitted by the \cf{addName} operation. The contract simply states our
high-level observation; for any effect $a$ which is also an \cf{AddName} effect
on the same object, it must be the case that $a$ is visible to $\cureff$ or
vice verse, or $a$ is $\cureff$. Here, $\sameobjZ$ and $\visZ$ are primitive
relations. This contract ensures that any two \cf{addName} operations will be
seen by each other. The contract for withdraw is similar to the
\cf{addNameCtrt}.

Since the deposit operation does not have any restrictions, its contract is
simply $\true$. Same is the case for \cf{addUser} operation. The contract for
get balance operation is:
\begin{smathpar}
\begin{array}{l}
\rsf{getBalCtrt} = \forall (a:\rsf{Deposit}), (b:\rsf{Withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \vee (\soZ \cap \sameobjZ) (a,\cureff) \Rightarrow \vis{a}{\cureff} \\
\qquad \vee (\soZ \cap \sameobjZ) (b,\cureff) \Rightarrow \vis{b}{\cureff}
\end{array}
\end{smathpar}

If a withdraw $b$ is visible to the get balance operation, then all deposit
operations $a$ visible to withdraw should be visible to the get balance
operation. This prevents negative balance anomalies. The rest of the contract
says that a get balance operation must witness previous deposit and withdraw
operations on the same object in the same session. This prevents missed update
anomalies.

Similar to contracts on operations, Quelea supports contracts on transactions.
The contract on the \cf{getBalanceName} transaction is given below:
\begin{smathpar}
\begin{array}{l}
\rsf{getBalanceNameCtrt} = \forall (a:\rsf{GetName}), (b:\rsf{GetBalance}), \\
\quad (c:\rsf{AddName}), (d:\rsf{AddUser}). ~\trans{a}{b}{c}{d} \wedge \so{a}{b} \\
\qquad \wedge ~\vis{c}{a} \wedge \sameobj{d}{b} \Rightarrow \vis{d}{b}
\end{array}
\end{smathpar}

$\trans{a}{b}{c}{d}$ says that the action pairs $a,b$ and $c,d$ are in the same
transaction, and the two transactions are distinct. The contract says
specifically forbids the anomaly presented in
Figure~\ref{fig:monotonicityAnomaly}. This is the desired semantics for the
\cf{getUserName} transaction.

The programmer simply defines such contracts on the operations and the
transactions. Quelea logically analyzes the contracts, and maps the
corresponding operation to the precise store consistency and isolation level.
Thus, Quelea equips the programmer with a declarative model for reasoning and
expressing eventually consistent programs.

>>>>

In this section, we motivate the need and benefit of declarative reasoning under
eventual consistency with the help of an example. We begin with a description of
our system and programming model, and introduce the primitive relations that
constitute an essential part of our contract language.

\subsection{System Model}
\label{sec:sysmod}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{Figures/SystemModel}
\caption{Quelea system model.}
\label{fig:sysmod}
\end{figure}

Figure~\ref{fig:sysmod} provides a schematic diagram of our system model. The
distributed store is composed of a collection of \emph{replicas}, each of which
stores a set of \emph{objects} ($x,y,\ldots$). We assume that every object is
replicated at every replica in the store. The state of an object at any replica
is the set of all updates (\emph{effects}) performed on the object. For
example, the state of $x$ at replica 1 is the set composed of effects $w^x_1$
and $w^x_2$.

Each object is associated with a set of \emph{operations}. The clients interact
with the store by invoking the operations on objects. The sequence of operations
invoked by a particular client on the store is called a \emph{session}. The
data store is typically accessed by a large number of clients (and hence
sessions) concurrently. Importantly, the clients are oblivious to which replica
an operation is applied to; the data store may choose to route the operation to
the closest replica for minimizing latency, or route it to a different replica
in order to balance the load in the system. For example, the operations $foo$
and $bar$ invoked by the same session on the same object, might end up being
applied to different replicas because replica 1 to which $foo$ was applied to
might be unreachable when invoking $bar$.

Suppose an operation $foo$ is invoked on a object $x$ with arguments $arg_1$ at
replica 1. The operation $foo$ simply \emph{reduces} over the current set of
effects at that replica on that object ($w^x_1$ and $w^x_2$), produces a result
$v1$ that is sent back to the client, and emits a \emph{single} new effect
$w^x_4$, which is appended to the state $x$ at replica 1. Thus, every operation
is evaluated over a \emph{snapshot} of the state of the object on which it is
invoked. In this case, we say that the effect $w^x_4$ \emph{witnessed} $w^x_1$
and $w^x_2$. Alternatively, $w^x_1$ and $w^x_2$ are \emph{visible} to $w^x_4$,
written logically as $\vis{w^x_1}{w^x_4} \wedge \vis{w^x_2}{w^x_4}$, where
$\visZ$ is the visibility relation between the effects. Visibility is an
irreflexive (an operation cannot see its own effect) and asymmetric relation,
and only relates effects produced by operations on the same object.

Read-only operation execution is similar except that they do not produce a new
effect. The effect added to a particular replica is asynchronously sent to
other replicas. We assume that \emph{eventually}, an effect applied at one
replica will be merged into all other replicas. The two effects $w^x_4$ and
$w^x_5$ that arise from the same session are said to be in \emph{session order}
(written logically as $\so{w^x_4}{w^x_5}$). Session order is a irreflexive,
transitive relation. The effects $w^x_4$ and $w^x_5$ arising from operations
applied to the same object $x$ are said to be under the \emph{same object}
relation, written $\sameobj{w^x_4}{w^x_5}$. Finally, we can associate every
effect with the operation that generated the effect with the help of a relation
$\operZ$. In the current example, $\oper{w^x_4}{foo}$ and $\oper{w^x_5}{bar}$
hold. For simplicity, we assume that all the operation names across all object
types are distinct.

The system model described allows all the inconsistencies associated with
eventual consistency. The goal of our work is to identify the precise
consistency level for each operation such that the application-level
constraints are not violated. In the next section, we will concretely describe
the challenges associated with constructing a consistent bank account on top of
eventually consistent data stores. Subsequently, we will illustrate how our
contract language, armed with the primitive relations $\visZ$, $\soZ$,
$\sameobjZ$ and $\operZ$, mitigates these challenges.

\subsection{Eventually Consistent Bank Account}

Suppose our goal is to implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance should never drop below 0. Let us assume that the bank account object
provides three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance},
with the assumption that the withdraw fails if the account has insufficient
balance to complete the withdraw operation. Our goal is to choose the correct
consistency level for each operation such that (1) the balance never drops
below zero and (2) the \cf{getBalance} operation never incorrectly show a
negative balance. To be able to do this, we need to consider the anomalies that
could arise under eventual consistency.

\begin{figure}
\centering
\subfigure[Unsafe Withdraw]{\label{fig:unsafeWithdrawAnomaly}\includegraphics[width=0.34\columnwidth]{Figures/Motivation4}}
\hfill
\subfigure[Negative Balance]{\label{fig:negativeBalanceAnomaly}\includegraphics[width=0.31\columnwidth]{Figures/Motivation2}}
\hfill
\subfigure[Missing update]{\label{fig:missingUpdateAnomaly}\includegraphics[width=0.26\columnwidth]{Figures/Motivation1}}
\caption{Anomalies possible under eventual consistency for the get balance operation.}
\label{fig:cleanliness_examples}
\end{figure}

Consider the execution shown in Figure~\ref{fig:unsafeWithdrawAnomaly}. Assume
that all operations in the figure are on the same bank account object with
initial balance as 0. Session 1 performs a \cf{deposit} of 100, followed by a
\cf{withdraw} of 80 in the same session. The \cf{withdraw} operation witnesses
the deposit and succeeds\footnote{Although visibility and session order
relations relate effects, we have abused the notation in these examples to
relate operations, with the idea that the relations relate the effect emitted
by those operations}. Subsequently, session 2 perform a \cf{withdraw}
operation, but importantly, due to eventual consistency, only witnesses the
\cf{deposit} from session 1, but not the subsequent withdraw. Hence, this
\cf{withdraw} also \emph{incorrectly} succeeds, violating the integrity
constraint. A subsequent \cf{getBalance} operation, that witnesses all the
previous operations and reports a negative balance.

It is easy to see that the only way to ensure that integrity constraint on
\cf{balance} is to prevent concurrent withdrawals on the same bank account
object. This can be done by insisting that \cf{withdraw} be executed with
strong consistency. Although this prevents the integrity constraint from being
violated, a strongly consistent \cf{withdraw} itself is not sufficient to
prevent \cf{getBalance} from ever reporting a negative balance to the end user.

Consider the execution shown in fig.~\ref{fig:negativeBalanceAnomaly}, which
consists of three concurrent sessions performing a \cf{deposit}, a
\cf{withdraw}, and a \cf{getBalance} operation, respectively, on the same bank
account object. Assume that initial balance in the account is zero. As the
\cf{vis} edge indicates, operation \cf{withdraw(50)} in session 2, can
witnesses the effects of \cf{deposit(100)} from session 1, leading it to
rightfully conclude that the current balance is 100. Consequently,
\cf{withdraw(50)} is performed safely. The effect of this \cf{withdraw}
operation subsequently becomes visible to the \cf{getBalance} in Session 3.
However, \cf{getBalance}, which only witnesses the effect of \cf{withdraw} from
session 2, but not the \cf{deposit} from session 1, reports the balance of
negative 50 to the user.

Negative balance is not the only inconsistency that the user is exposed to
under an eventually consistent bank account. Figure
~\ref{fig:missingUpdateAnomaly} shows an execution where the \cf{getBalance}
operation in a session does not witness the effects of \cf{withdraw} operation
performed in the same session, possibly because \cf{getBalance} was served by a
replica that has not yet merged the \cf{withdraw} effect, served by a different
replica. This anomaly leads the user to incorrectly conclude that the
\cf{withdraw} operation failed to go through.

Although it is easy to understand the reasons behind the occurrence of the
aforementioned negative balance and missing update anomalies, the fix is not
quite apparent. Making \cf{getBalance} a strongly consistent operation is
definitely sufficient to avert anomalies, but is it really necessary? Given
that a strongly consistent operation is significantly expensive, both in terms
of availability and monetary overheads~\cite{DynamoDB,Pileus}, we ideally would
not want to request strong consistency, unless there are no alternatives
available. Unfortunately, this requires us to understand subtle differences in
semantics of various kinds of weak consistency alternatives, offered by the
data stores.

@@

\name automates this error-prone task by letting the programmer declare
application-level consistency constraints as \emph{contracts}. The key
intuition is that, in the case of the bank account example, any execution that
does not exhibit the anomalies described in the previous section is a
\emph{well-formed} execution on the bank account object. In the same vein as
axiomatic memory models for shared memory multicore
processors~\cite{Demange2013,Sarkar2011}, the programmer axiomatically
describes the set of legal executions on top of each replicated date type over
a trace of operation invocations on that type. Quelea ensures that any
execution on top of such datatypes always belongs to the set of well-formed
executions on that type.

In the bank account example, we observed that, in order to preserve the
integrity constraint, the \cf{withdraw} operation must be strongly consistent.
That is, given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible
to $b$ or vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract over \cf{withdraw}:
\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}
In the above contract, $\cureff$ stands for the effect emitted by the
\cf{withdraw} operation. The syntax $a : \rcf{withdraw}$ states that $a$ is an
effect which was emitted by a \cf{withdraw} operation i.e)
$\oper{a}{\rcf{withdraw}}$ holds. The contract says that, if the current
operation emits an effect $\cureff$, then for any operation $a$ which was
emitted by a \cf{withdraw} operation, it is the case that $a = \cureff$ or $a$
is visible to $\cureff$ or vice versa. Any execution on a bank account object
that preserves the above contract for a \cf{withdraw} operation is said to have
a correct implementation of \cf{withdraw}.

For the \cf{getBalance}, we construct a contract  precisely to prevent the
anomalies described previously. The contract is shown below:
\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \wedge~ (\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff}
\end{array}
\end{smathpar}

If a \cf{withdraw} $b$ is visible to the get balance operation, then all
\cf{deposit} operations $a$ visible to the \cf{withdraw} should be visible to
the get balance operation. This prevents negative balance anomalies. The syntax
$c : \rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an effect that was
emitted either by a \cf{deposit} operation or by a \cf{withdraw} operation. The
contract language provides relational operators to compose relations. The
syntax $(R_1 \cap R_2)(a,b)$ is equivalent to $R_1(a,b) \wedge R_2(a,b)$. The
last line of the above contract says that if a \cf{deposit} or a \cf{withdraw}
precedes a \cf{getBalance} operation in session order, and is applied on the
same object as the \cf{getBalance} operation, then it must be the case that the
\cf{getBalance} operation witnesses the effects of the preceding operations.
Since the deposit operation does not have any restrictions, its contract is
simply $\true$.

\subsection{From contracts to implementation}

Notice that the contracts for \cf{withdraw} and \cf{getBalance} only express
the application-level consistency requirements, and make no reference to the
semantics of the underlying store. To write such contracts, a programmer only
needs to reason about the semantics of the application under the \name system
model. The mapping of application-level consistency requirements to appropriate
store-level guarantees is done automatically behind the screen.

How might one go about ensuring that any execution adheres to the contract? The
challenge is that contract is a declarative characterization of correctness of
execution, while what we need is an operational procedure for \emph{enforcing}
the contracts. One strategy is to adopt speculative execution as espoused by
other eventually consistent systems, where operations are tentatively applied
as they are received from the client or other replicas. We can maintain a
runtime manifestation of the axiomatic characterization and check the
well-formedness at runtime, and rollback the execution if
ill-formed~\cite{rxcml}. However, the overhead of state maintenance and the
complexity of user-defined contracts (unlike simple tests such as cyclic
dependence) make this technique unviable.

Instead, Quelea statically analyzes the contracts using an SMT solver, and maps
each operation to a particular store-level consistency under which the contract
would be preserved. In particular, the key idea is that the system implementor
axiomatically describes the store semantics in the same contract language as
the one used to express the contract on the operation. We translate the
contracts down to decidable subset of first-order logic~\cite{}, and discharge
proof obligation of checking whether a particular store consistency level is
sufficient for the given contract with the help of an SMT solver. The next
section describes the contract language and classification.
